---
title: "R and RStudio Basics for Complete Beginners"
author: "Nurlan Kerimov"
output: html_document
---

# Welcome

This lesson is for students who are completely new to R, RStudio, and programming.

You will learn the basics step by step:

- R Markdown document structure
- Variables
- Running code in the correct order
- Data types
- Arithmetic operations
- if/else conditions
- Iterators (`for`, `while`)
- Functions

## How to run code in RStudio

In this file, code appears inside **code chunks** (the gray boxes).

- Run one line or selected lines: select code and press `Cmd + Enter` (Mac) or `Ctrl + Enter` (Windows).
- Run the current chunk: click the green **Run/Play** button on the top-right of the chunk.
- Run all chunks: use the **Run** menu in RStudio.

Execution order matters. If line 2 uses a variable from line 1, you must run line 1 first.

---

# 1) R Markdown document structure

An `.Rmd` file usually has:

1. A YAML header at the top (between `---` and `---`)
2. Markdown text for explanation
3. R code chunks for executable code

Example chunk:

```{r}
message("Hello from a code chunk")
```

Try it yourself: click the green **Run** button in the chunk above.

---

# 2) Variables

A **variable** stores a value in memory.
In R, we usually assign values with `<-`.

```{r}
name <- "Aylin"
age <- 22
is_student <- TRUE

name
age
is_student
```

Try it yourself: change `name` and `age`, then run the chunk again.

---

# 3) Running code line by line (order is important)

This example will fail if you run only the last line first.

```{r}
first_number <- 10
second_number <- 5
total <- first_number + second_number

total
```

If you run only `total` before defining `first_number` and `second_number`, R will show an error.

Use the green **Run** button to run the whole chunk when you are not sure whether earlier lines were executed.

---

# 4) Data types

Common beginner data types in R:

- Character (text): `"hello"`
- Numeric (number): `42` or `3.14`
- Logical (boolean): `TRUE` or `FALSE`

```{r}
city <- "Baku"       # character
temperature <- 18.5  # numeric
is_raining <- FALSE  # logical

class(city)
class(temperature)
class(is_raining)
```

Try it yourself: create one more variable for each type.

---

# 5) Arithmetic operations

Basic arithmetic operators:

- `+` add
- `-` subtract
- `*` multiply
- `/` divide
- `^` power

```{r}
x <- 12
y <- 3

x + y
x - y
x * y
x / y
x ^ y
```

Try it yourself: change `x` and `y` and predict outputs before running.

---

# 6) if/else conditions

Use `if` and `else` to make decisions.

```{r}
score <- 72

if (score >= 60) {
  message("Pass")
} else {
  message("Fail")
}
```

Try it yourself: set `score <- 45` and run again.

---

# 7) Iterators: for and while

## for loop

A `for` loop repeats code for each value in a sequence.

```{r}
for (i in 1:5) {
  print(i)
}
```

## while loop

A `while` loop runs while a condition stays true.

```{r}
counter <- 1

while (counter <= 3) {
  print(counter)
  counter <- counter + 1
}
```

Try it yourself: change loop limits and run again.

## foreach-style in R (using `lapply`)

In many languages, people say "foreach". In base R, a beginner-friendly equivalent is often `lapply`.

```{r}
numbers <- c(2, 4, 6)

lapply(numbers, function(n) {
  n * 10
})
```

This applies the same function to each element (similar to a foreach idea).

---

# 8) Functions

A **function** is reusable code that can take inputs and return outputs.

```{r}
add_tax <- function(price, tax_rate) {
  final_price <- price * (1 + tax_rate)
  return(final_price)
}

add_tax(100, 0.18)
```

Try it yourself: call `add_tax(250, 0.10)`.

---

# Practice mini-task

Complete this short exercise:

1. Create a variable `daily_hours` with a number.
2. Use `if/else` to print `"Busy day"` if `daily_hours >= 8`, otherwise `"Light day"`.
3. Use a `for` loop to print numbers from 1 to `daily_hours`.
4. Write a function `double_value()` that returns input multiplied by 2.

```{r}
# 1
daily_hours <- 6

# 2
if (daily_hours >= 8) {
  message("Busy day")
} else {
  message("Light day")
}

# 3
for (n in 1:daily_hours) {
  print(n)
}

# 4
double_value <- function(x) {
  x * 2
}

double_value(daily_hours)
```

---

# Recap

You practiced:

- R Markdown structure
- Variables and data types
- Arithmetic
- Code execution order
- if/else logic
- Loops (`for`, `while`, and foreach-style with `lapply`)
- Functions

If something does not work, first run earlier lines or run the full chunk with the green **Run/Play** button.
